Implementation of Parallel Algorithms for ARIMA in Distributed Database Systems

### 2013-10-04 draft.tex
Minor changes
Added figures and the required format style files

\documentclass[english,12pt]{article}
\usepackage{babel}
\usepackage{natbib}
\usepackage[latin1]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm2e}

%\makeatletter
\linespread{1.25}

% \pagestyle{headings}

@@

@@

Time series analyses are important in econometrics, mathematical finance, weather forecasting, earthquake prediction and many other fields.
For example, one of the most conspicuous data analytics task -- stock price forecasting -- falls into the category of time series analysis.

However, different from other common data analytics tasks, time series data have a natural temporal ordering.
And many time series modeling methods, such as ARIMA~\cite{arima} and Cox proportional hazards~\cite{cox} therefore depend on sequential processing of time series data, which raises a challenge for the data-parallel implementation in MADlib.

In MADlib, we attack this challenge using an idea illustrated below [TODO: add a simple figure to compare MADlib ARIMA and order aggregate with iterations].

@@

The MADlib project was initiated in late 2010. Currently MADlib's latest
version is 1.2. By itself MADlib provides a pure SQL interface. An R front-end
named PivotalR~\cite{pivotalr} is also created by the Predictive Analytics Team
at Pivotal Inc. A Python wrapper~\cite{python-madlib} for MADlib is also
available.

@@

@@

In the next few subsections, we will first describe the algorithm that we have
used to solve the problem of maximization of partial log-likelihood. Then, we
point out the reason why it is difficult to make the algorithm run parallel.
The difficulty is not restricted to this specific algorithm. It is a
common situation in all algorithms that fit ARIMA models. Actually it is easy
to see that many machine learning models have similar problems. Next, we
describe our solution to this problem. Our method is quite general and can
parallelize similar problems. Then we describe a simple method to improve the
performance of our algorithm. In the end, we discuss the possible ways to
generalize our methods to more algorithms.

@@

\begin{algorithm}[ht]
\KwIn{An initial guess for parameters $\vec{\phi}_0, \vec{\theta}_0, \mu_0$}
\KwOut{The parameters that maximize the likelihood $\vec{\phi}^*,
  \vec{\theta}^*, \mu^*$}
    $k \leftarrow 0$; $v \leftarrow 2$;
    $(\vec{\phi},\vec{\theta},\mu) \leftarrow
    (\vec{\phi}_0,\vec{\theta}_0,\mu_0)$\;
    Calculate $Z(\vec{\phi},\vec{\theta},\mu)$ with
    Eq. (9).  \textit{\footnotesize \# Vector of errors}\;
    $A \leftarrow J^T J$   \textit{\footnotesize \# The Gauss-Newton Hessian
    approximation}\;
    $u \leftarrow \tau * \max_i(A_{ii})$ \textit{\footnotesize \# Weight of the
    gradient-descent step}\;
    $g \leftarrow J^T Z(\vec{\phi},\vec{\theta},\mu)$ \textit{\footnotesize \#
    The gradient descent step}\;
	$ \text{stop} \leftarrow (\|g\|_{\infty} \le \epsilon_1)$
    \textit{\footnotesize \# Termination Variable}\;
	\While{not stop and $k < k_{\max}$}{
		$k \leftarrow k + 1$\;
		\Repeat{stop or $\rho > 0$}{
			$\delta \leftarrow {(A + u \times \text{diag}(A))}^{-1} g$
            \textit{\footnotesize \# Calculate step direction}\;
            \eIf(\textit{\footnotesize \# Change  is too small
              to
              continue.})
            {$\| \delta \| \le \epsilon_2 \|
              (\vec{\phi},\vec{\theta},\mu) \|$} {
				$\text{stop} \leftarrow \text{true}$\;
			}{
				$(\vec{\phi}_{new},\vec{\theta}_{new},\mu_{new})
                \leftarrow (\vec{\phi},\vec{\theta},\mu) + \delta$
                \textit{\footnotesize \# A trial step}\;
				$\rho \leftarrow (\|
                Z(\vec{\phi},\vec{\theta},\mu)\|^2 - \|
                Z(\vec{\phi}_{new},\vec{\theta}_{new},\mu_{new})\|^2
                )/(\delta^T(u \delta + g))$ \;
                \eIf(\textit{\footnotesize \# Trial step was good}){$\rho >
                    0$}{
					$(\vec{\phi},\vec{\theta},\mu) \leftarrow
                    (\vec{\phi}_{new},\vec{\theta}_{new},\mu_{new})$
                    \textit{\footnotesize \# Update variables}\;
                    Calculate $Z(\vec{\phi},\vec{\theta},\mu)$ with
                    Eq. (9); $A \leftarrow J^T J$; $g \leftarrow J^T
                    Z(\vec{\phi},\vec{\theta},\mu)$\;
					$ \text{stop} \leftarrow (\|g\|_{\infty} \le
                    \epsilon_1)$ or $(\|
                    Z{(\vec{\phi},\vec{\theta},\mu)}^2 \| \le
                    \epsilon_3)$ \;
                    $v \leftarrow 2$;
					$u \rightarrow u * \max(1/3, 1 - {(2\rho - 1)}^3 )$\;
                } (\textit{\footnotesize \# Trial step was bad}){
                    $v \leftarrow 2 v$;
					$u \leftarrow u v$\;
				}
			}
		 }
	}
	$(\vec{\phi}^*,\vec{\theta}^*,\mu^*) \leftarrow (\vec{\phi},\vec{\theta},\mu)$\;

@@

It is very easy to see that Eqs.
(\ref{eq:error-terms},~\ref{eq:J1},~\ref{eq:J2},~\ref{eq:J3}) are
difficult to parallelize. Each step computation uses the result from the
previous step. Therefore, we have to scan through the data sequentially to
compute the quatities in these equations.

@@

>>>>

\documentclass[english,12pt]{article}

\usepackage{babel}

\usepackage{natbib}

\usepackage[latin1]{inputenc}

\usepackage[T1]{fontenc}

\usepackage{amsmath}

\usepackage{amssymb}

\usepackage{algorithm2e}

\usepackage{graphicx}

\makeatletter
\linespread{1.25}

\pagestyle{headings}

@@

{\bf keywords}: distributiv, parallel computation, database, machine
learning, big data, ARIMA

@@

Time series analyses are important in econometrics, mathematical
finance, weather forecasting, earthquake prediction and many other
fields. For example, one of the most conspicuous data analytics task,
stock price forecasting, falls into the category of time series
analysis.

However, different from other common data analytics tasks, time series
data have a natural temporal ordering. And many time series modeling
methods, such as ARIMA~\cite{arima} and Cox proportional
hazards~\cite{cox}, therefore depend on sequential processing of time
series data, which raises a challenge for the data-parallel
implementation in MADlib.

In MADlib, we attack this challenge using an idea illustrated below
[TODO add a simple figure to compare MADlib ARIMA and order aggregate
with iterations].

@@

The MADlib project was initiated in late 2010. Currently MADlib's latest
version is 1.3. By itself MADlib provides a pure SQL interface. An R
front-end named PivotalR~\cite{pivotalr} is also created by the
Predictive Analytics Team at Pivotal Inc.

@@

Although MADlib does not do parallel computation on PostgreSQL, it is
still valuable for process big data sets that cannot be loaded into
memory. For example, in this paper we describe our implementation of
ARIMA in MADlib. We compare our implementation

@@

In the next few subsections, we will first describe the algorithm that
we have used to solve the problem of maximization of partial
log-likelihood. Then, we point out the reason why it is difficult to
make the algorithm run parallel.  The difficulty is not restricted to
this specific algorithm. It is a common situation in all algorithms
that fit ARIMA models. Actually it is easy to see that many machine
learning models have similar problems. Next, we describe our solution
to this problem. Our method is quite general and can parallelize
similar problems. Then we describe a simple method to improve the
performance of our algorithm. In the end, we discuss the possible ways
to generalize our methods to other algorithms.

@@

\begin{algorithm}[ht]
\KwIn{An initial guess for parameters $\vec{\phi}_0, \vec{\theta}_0, \mu_0$}
\KwOut{The parameters that maximize the likelihood $\vec{\phi}^*,
  \vec{\theta}^*, \mu^*$}
    $k \leftarrow 0$; $v \leftarrow 2$;
    $(\vec{\phi},\vec{\theta},\mu) \leftarrow
    (\vec{\phi}_0,\vec{\theta}_0,\mu_0)$\;
    Calculate $Z(\vec{\phi},\vec{\theta},\mu)$ with
    Eq. (9).  \textit{\footnotesize \# Vector of errors}\;
    $A \leftarrow J^T J$   \textit{\footnotesize \# The Gauss-Newton Hessian
    approximation}\;
    $u \leftarrow \tau * \max_i(A_{ii})$ \textit{\footnotesize \# Weight of the
    gradient-descent step}\;
    $g \leftarrow J^T Z(\vec{\phi},\vec{\theta},\mu)$ \textit{\footnotesize \#
    The gradient descent step}\;
	$ \text{stop} \leftarrow (\|g\|_{\infty} \le \epsilon_1)$
    \textit{\footnotesize \# Termination Variable}\;
	\While{not stop and $k < k_{\max}$}{
		$k \leftarrow k + 1$\;
		\Repeat{stop or $\rho > 0$}{
            $\delta \leftarrow{(A + u \times \text{diag}(A))}^{-1} g$
            \textit{\footnotesize \# Calculate step direction}\;
            \eIf(\textit{\footnotesize \# Change  is too small
              to
              continue.})
            {$\| \delta \| \le \epsilon_2 \|
              (\vec{\phi},\vec{\theta},\mu) \|$} {
				$\text{stop} \leftarrow \text{true}$\;
			}{
				$(\vec{\phi}_{new},\vec{\theta}_{new},\mu_{new})
                \leftarrow (\vec{\phi},\vec{\theta},\mu) + \delta$
                \textit{\footnotesize \# A trial step}\;
				$\rho \leftarrow (\|
                Z(\vec{\phi},\vec{\theta},\mu)\|^2 - \|
                Z(\vec{\phi}_{new},\vec{\theta}_{new},\mu_{new})\|^2
                )/(\delta^T(u \delta + g))$ \;
                \eIf(\textit{\footnotesize \# Trial step was good}){$\rho >
                    0$}{
					$(\vec{\phi},\vec{\theta},\mu) \leftarrow
                    (\vec{\phi}_{new},\vec{\theta}_{new},\mu_{new})$
                    \textit{\footnotesize \# Update variables}\;
                    Calculate $Z(\vec{\phi},\vec{\theta},\mu)$ with
                    Eq. (9); $A \leftarrow J^T J$; $g \leftarrow J^T
                    Z(\vec{\phi},\vec{\theta},\mu)$\;
					$ \text{stop} \leftarrow (\|g\|_{\infty} \le
                    \epsilon_1)$ or $(\|
                    Z{(\vec{\phi},\vec{\theta},\mu)}^2 \| \le
                    \epsilon_3)$ \;
                    $v \leftarrow 2$;
					$u \rightarrow u * \max(1/3, 1 - {(2\rho - 1)}^3 )$\;
                } (\textit{\footnotesize \# Trial step was bad}){
                    $v \leftarrow 2 v$;
					$u \leftarrow u v$\;
				}
			}
		 }
	}
	$(\vec{\phi}^*,\vec{\theta}^*,\mu^*) \leftarrow (\vec{\phi},\vec{\theta},\mu)$\;

@@

It is very easy to see that Eqs.
(\ref{eq:error-terms},~\ref{eq:J1},~\ref{eq:J2},~\ref{eq:J3}) are
difficult to parallelize. Each step computation uses the result from
the previous step. Therefore, we have to scan through the data
sequentially to compute the quatities in these equations. If the data
set is very large then it

@@

In\ order to utilize the data parallel capability of MPP database, we
propose to segment the whole time series data into a set of
consecutive time series data, numbering from $1$ to $N$, each subset
comprising a sequence of consecutive time series data of size
$K$. Since the model fitting involves multile iterations of
computations, for any subset $i$ execpt for the first one, we use the
results of the $(i-1)$-th subset from the previous iteration as the
initial values. The first subset's initial values are known to Thus in
each iteration, the computation for each subset of data does not need
to wait for the results from the previous subset. In this way, the
model fitting computaion can be parallelized.

Besides, we find that aggregating each subset of consecutive time
series data into an array (i.e.\ a data chunk) can greatly simplify
the implementation (for example, from a stateful window function
implementation to a plain UDF implementation) and accelerate the
computation (mainly due to the reduction of I/O overhead and function
call overhead). Furthermore, we can specify the value of size $K$
according to the system configuration so that each chunk of data can
be completely loaded into the avaialble
memory. Fig. (\ref{fig:seq_vs_dist}) shows the different designs of
using a sequential implementation and a parallel implementation.

\begin{figure}[ht]
  \begin{centering}
    \includegraphics[scale=0.25]{sequential_vs_distributed.png}
  \end{centering}
  \caption{\label{fig:seq_vs_dist} The comparison between sequential
    and distributed designs. }
\end{figure}

The following SQL shows how we segment and redistribute the data according to the above proposals.

\begin{verbatim}
create temp table dist_table as
  select
    ((tid - 1) / chunk_size)::integer + 1 as distid, 
    tid, tval
  from 
      input_table
distributed by (distid)
\end{verbatim}

The chunk size has some effect on the performance, but the effect is
not very big.

\begin{figure}[ht]
  \begin{centering}
    \includegraphics[angle=-90,scale=0.4]{1e8_time_vs_chunksize}
  \end{centering}
  \caption{\label{fig:chunksize_effect} We fit a time series with
    length of $10^8$ using different chunk sizes. The execution times
    are around $400$ $\sim$ $500$ secs, and are quite stable. }
\end{figure}

\begin{figure}[ht]
  \begin{centering}
    \includegraphics[angle=-90,scale=0.4]{gp_time_vs_size}
  \end{centering}
  \caption{\label{fig:gp_time_size} We fit a time series with length
    of $10^8$ using different chunk sizes. The execution times are
    around $400$ $\sim$ $500$ secs, and are quite stable. }
\end{figure}

\begin{figure}[ht]
  \begin{centering}
    \includegraphics[angle=-90,scale=0.4]{pg_time_vs_size}
  \end{centering}
  \caption{\label{fig:gp_time_size} We fit a time series with length
    of $10^8$ using different chunk sizes. The execution times are
    around $400$ $\sim$ $500$ secs, and are quite stable. }
\end{figure}
